Racial Force Probe

### 2024-02-15 paper/paper.qmd
2 graphs added and qmd structure improved

---
title: "Racial Force Probe"
subtitle: "The study looks on a single scientific argument about whether or not certain police officers discriminate against people based on their race. "
author:
  - Monica Sainani
  - Kaavya Kalani
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "This research endeavors to scrutinize the well-being of individuals within the prison system, focusing on the prevalence of physical diseases, mental health issues, and mortality rates among incarcerated populations. The prison environment presents unique challenges that can significantly impact the overall health of individuals, making it imperative to investigate the multifaceted dimensions of prisoners' well-being. Utilizing data collected over the past [insert number of years], this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners. By examining the intersection of physical and mental health alongside mortality rates, the research aims to contribute valuable insights into the current state of prisoners' well-being. This analysis not only serves to inform our understanding of the challenges faced by incarcerated individuals but also provides a foundation for evidence-based policy recommendations aimed at fostering improved conditions within correctional facilities. Through this exploration, we strive to contribute to the ongoing dialogue surrounding the intersection of criminal justice and public health, emphasizing the importance of compassionate and effective strategies for promoting the well-being of those within the prison system."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(palmerpenguins)
```

# Introduction
Health and Safety of Prisoners Today

@@

# Data {#sec-data}

# Model

@@

@@

# Appendix {-}

# Additional data details

>>>>

---
title: "Racial Force Probe"
author:
  - Monica Sainani
  - Kaavya Kalani
  - Aamishi Avarsekar
thanks: "Code and data are available at: https://github.com/kaavyakalani26/prison_health.git and replication aspects are available at: https://www.socialsciencereproduction.org/reproductions/82b709e2-4aa0-40e8-8abe-195c51ae739b/index"
date: today
date-format: long
abstract: "This research endeavors to scrutinize the well-being of individuals within the prison system, focusing on the prevalence of physical diseases, mental health issues, and mortality rates among incarcerated populations. The prison environment presents unique challenges that can significantly impact the overall health of individuals, making it imperative to investigate the multifaceted dimensions of prisoners' well-being. Utilizing data collected over the past [insert number of years], this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners. By examining the intersection of physical and mental health alongside mortality rates, the research aims to contribute valuable insights into the current state of prisoners' well-being. This analysis not only serves to inform our understanding of the challenges faced by incarcerated individuals but also provides a foundation for evidence-based policy recommendations aimed at fostering improved conditions within correctional facilities. Through this exploration, we strive to contribute to the ongoing dialogue surrounding the intersection of criminal justice and public health, emphasizing the importance of compassionate and effective strategies for promoting the well-being of those within the prison system."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

#### Preamble ####
# Purpose: Read the cleaned dataset to use it for the analysis
# Author: Kaavya Kalani, Monica Sainani, Aamishi Avarsekar
# Date: 14 February 2024
# Contact: kaavya.kalani@mail.utoronto.ca, monica.sainani@mail.utoronto.ca, aamishi.avarsekar@mail.utoronto.ca
# License: MIT
# Pre-requisites: run scripts/01-download_data.R followed by scripts/02-data_cleaning.R

#### Workspace setup ####
# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("here")
# install.packages("ggplot2")
# install.packages("janitor")

library(ggplot2)
library(tidyverse)
library(knitr)
library(here)
library(janitor)
```

# Introduction

Health and Safety of Prisoners Today

@@

# Data

@@

```{r}
#| message: false
#| echo: false
#| label: fig-overall-death
#| fig-cap: Causes of deaths among the prisoners
```

```{r}
#| message: false
#| echo: false
#| label: fig-covid
#| fig-cap: Covid trends in the jail
```

```{r}
#| message: false
#| echo: false
#| label: fig-death-vs-gender
#| fig-cap: Causes of deaths among the prisoners by gender

health_data_count_by_demographics <- read_csv('data/msfp0116stt09data.csv')

health_demographic_data_clean <-
  clean_names(health_data_count_by_demographics) |>
  select("x1", "cancer", "heart_disease", "liver_disease",
         "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
         "drug_alcohol_intoxication", "accident", "homicide_d")

# Create a mapping for all causes
cause_mapping <- c(
  "cancer" = "Cancer",
  "heart_disease" = "Heart Disease",
  "liver_disease" = "Liver Disease",
  "respiratory_disease" = "Respiratory Disease",
  "aids_related" = "Aid related",
  "all_other_illnesses" = "All other illnesses",
  "suicide" = "Suicide",
  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
  "accident" = "Accident",
  "homicide_d" = "Homicide"
)

# Rename the causes in the dataset
health_demographic_data_clean <- health_demographic_data_clean %>%
  rename_with(~ cause_mapping[.], -x1)

gender_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("Male", "Female"))

gender_health_data_long <- pivot_longer(gender_health_data, cols = -x1, names_to = "Cause", values_to = "Count")

# Calculate proportions within each category
gender_health_data_long <- gender_health_data_long %>%
  group_by(x1) %>%
  mutate(Proportion = Count / sum(Count))

# Plot the proportions graph using geom_bar with position = "fill"
ggplot(gender_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
  labs(title = "Proportions of Causes of Death by Category",
       x = "Gender",
       y = "Proportion") +
  scale_fill_discrete(labels = cause_mapping) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| message: false
#| echo: false
#| label: fig-death-vs-race
#| fig-cap: Causes of deaths among the prisoners by race

# Read data
health_data_count_by_demographics <- read_csv('data/msfp0116stt09data.csv')

# Clean and organize data
health_demographic_data_clean <-
  clean_names(health_data_count_by_demographics) |>
  select("x1", "cancer", "heart_disease", "liver_disease",
         "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
         "drug_alcohol_intoxication", "accident", "homicide_d")

# Pivot data to long format
health_demographic_data_long <- pivot_longer(health_demographic_data_clean, cols = -x1, names_to = "Cause", values_to = "Count")

# Filter data for specific demographics (e.g., "Male" and "Female")
filtered_df <- health_demographic_data_clean[health_demographic_data_clean$x1 %in% c("Male", "Female"),]

# Filter data for specific races (e.g., "White", "Black", "Hispanic", "Other")
race_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("White", "Black", "Hispanic", "Other"))

# Pivot data to long format for race
race_health_data_long <- pivot_longer(race_health_data, cols = -x1, names_to = "Cause", values_to = "Count")

# Calculate proportions instead of counts
race_health_data_long <- race_health_data_long %>%
  group_by(x1) %>%
  mutate(Proportion = Count / sum(Count))

# Create a mapping for all causes
cause_mapping <- c(
  "cancer" = "Cancer",
  "heart_disease" = "Heart Disease",
  "liver_disease" = "Liver Disease",
  "respiratory_disease" = "Respiratory Disease",
  "aids_related" = "Aid related",
  "all_other_illnesses" = "All other illnesses",
  "suicide" = "Suicide",
  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
  "accident" = "Accident",
  "homicide_d" = "Homicide"
)

# Rename the causes in the dataset
race_health_data_long <- race_health_data_long %>%
  mutate(Cause = cause_mapping[as.character(Cause)])

# Create a bar plot showing proportions with black borders
ggplot(race_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
  labs(title = "Proportions of Causes of Death by Category",
       x = "Category",
       y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

@@

# Appendix
